Adaption of the Chumbley Score to matching of bullet striation marks

### 2017-10-18 doc/paper.Rmd
references added
comments and some updates to bib info

---
title: "Adaption of the Chumbley Score to matching of bullet striation marks"
authors:
- affiliation: Department of YYY, University of XXX
  name: Ganesh Krishnan
  thanks: The authors gratefully acknowledge ...
- affiliation: Department of ZZZ, University of WWW
  name: Heike Hofmann
biblio-style: apsr
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- 3 to 6 keywords
- that do not appear in the title
bibliography: bibliography
abstract: null
---

@@

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "",
  fig.height = 6,
  fig.width = 6,
  fig.align = "center",
  out.width= '\\textwidth',
  cache = FALSE,
  fig.path='figures/',
  echo=FALSE
)
options(knitr.table.format = "latex")
library(tidyverse)
```

@@

```{r data}
if (!file.exists("../all-sigs.rds")) {
  files <- dir("../data/signatures", pattern="csv")
  all <- data.frame()
  for (file in files) {
    tmp <- read.csv(file=file.path(path="../data/signatures", file)) 
    if (length(grep("x",names(tmp)) > 0)) tmp <- tmp %>% select(-x)
    all <- rbind(all, tmp)
  }
  all <- all %>% filter(land1_id < land2_id)
  saveRDS(all, file="../data/all-sigs.rds")
} else {
  all <- readRDS("../all-sigs.rds")
}

@@

## Problem statement: same source matching based on striations

\hh{'considered to be unique' by AFTE - that is right, but we should not use this argument, and I don't think we have to. It's better to focus on the problem at hand: striation marks are used to determine same source for bullets. XXX we could include some figures here XXX Current standards ask for a visual inspection of the marks under a comparison microscope by a firearms examiner. One of the problems raised by the PCAST report in 2009 is the subjectivity involved in this comparison. There are different approaches used to achieve more objectivity and assess error rates. XXX here we need some examples of approaches XXX
For toolmarks, the Chumbley score has been suggested to determine same-source. XXX summary of results from original paper and the adjustment XXX
In this paper we are interested in how this method has to be adjusted to best be applied to the smaller marks on bullets. }

@@

Compairing pairs of toolmarks with the intention of matching it to a tool has been studied in the past, and \citet{chumbley} have described in their paper an algorithm and analytic method that compares two toolmarks and comes to the conclusion if they are from the same tool. The motivation for this lies in the idea of developing analytical methods for which error rates can be found. This moves reduces subject bias and empirical based studies to designate two toolmarks as matches or non-matches, although \citet{chumbley} use this kind of a setup to validate their proposed analytic and quantitative algorithm.

The data that is used by \citet{chumbley} is generated by a surface profileometer that gives the height in terms of distance along a linear trace. This is taken perpendicular to the striations present in the toolmark. Two such trace are then compared using the algorithm.

Th algoritm works in two phases, namely, an optimization step and a validation step, at the end of which a Mann Whitney U statistic is calculated. \citet{hadler} in their paper proposed an improvement to this algorithm by trying to remove mutual dependence of parameters (due to serial correlation in surface depth values of a toolmark and because of a random sampling sub step in the validation phase which makes a group of pixels to be chosen more than once and hence introduces lack of independence) in certain steps, especially because the Mann-Whitney U statistic that is later calculated in the algorithm and used as a measure to differentiate between matches and non-matches works under the assumption of independence of parameters.

@@

A maximum correlation statistic is used to identify the region of best agreement, with the maximum usually seen being near 1 for both cases of matches ( which is intuitively what we expect) and non matches (which is not what we intuitively expect)

First, there are a very large number of correlations calculated, for eg if the window of optimization was defined as 200 and toolmark pixel length is around 1000 then we have 1200-(200-1) = 1001. Can be thought of ordered arrangements without replacement where n = 1001 and r = 1.

@@

This introduces a problem as if we go too small in the window of optimization the unique feature of the trace segments are lost and seem similar, and too large sizes vastly reduces the weightage of small features that would otherwise uniquely classify a signature and hence identify the region of agreement.

@@

@@

```{r win-comparison, fig.width=8, fig.height=5, out.width='\\textwidth', fig.cap=" The two plots on the left show how the same shift behaves in case of a matching pair and the two plots on the right show how the different shift behaves in case of a matching pair"}
a<- matrix(NA, ncol = 3, nrow = 2)
toolmark.falsepositive<- data.frame(a)
colnames(toolmark.falsepositive)<- c("Classification", "Match", "Non-Match")
toolmark.falsepositive$Classification<- c("Match", "Non-Match")
toolmark.falsepositive$Match<- c(47,0)
toolmark.falsepositive$`Non-Match`<- c(3,50)
toolmark.falsepositive

@@

```{r type2, fig.width=8, fig.height=5, out.width='\\textwidth', fig.cap="Type II error rates observed across a range of window sizes for optimization $wo$. For a window size of $wo = 120$ we see a drop in type II error rate across all type I rates considered. Smaller validation sizes $wv$ are typically associated with a smaller type II error."}
errors %>% filter(wv %in% c(30,50)) %>%
  ggplot(aes(x = wo, y = beta, colour=factor(alpha))) + 
  geom_point(aes(shape=factor(wv))) +
#  facet_grid(.~wo, labeller="label_both") + 
  geom_smooth(aes(group=alpha), se=FALSE, method="loess") +
  theme_bw() +
  scale_colour_brewer(expression("Nominal type I error"~alpha), palette="Set2") +
  scale_shape_discrete("Size of validation\nwindow, wv") +
  xlab(expression("Window size for optimization, "~wo)) +
  ylab("Type II error rate")

>>>>

---
title: "Adaption of the Chumbley Score to matching of bullet striation marks"
authors:
- affiliation: Department of Statistics, Iowa State University 
  name: Ganesh Krishnan
  thanks: The authors gratefully acknowledge ...
- affiliation: Department of Statistics and csafe, Iowa State University 
  name: Heike Hofmann
biblio-style: apsr
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- 3 to 6 keywords
- that do not appear in the title
bibliography: bibliography
abstract: null
---

@@

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "",
  fig.height = 6,
  fig.width = 6,
  fig.align = "center",
  out.width= '\\textwidth',
  cache = FALSE,
  fig.path='figures/',
  echo=FALSE,
  cache=TRUE
)
options(knitr.table.format = "latex")
library(tidyverse)
```

@@

```{r data}
if (!file.exists("../data/all-sigs.rds")) {
  files <- dir("../data/signatures", pattern="csv")
  all <- data.frame()
  for (file in files) {
    tmp <- read.csv(file=file.path(path="../data/signatures", file)) 
    if (length(grep("x",names(tmp)) > 0)) tmp <- tmp %>% select(-x)
    all <- rbind(all, tmp)
  }
  all <- all %>% filter(land1_id < land2_id)
  saveRDS(all, file="../data/all-sigs.rds")
} else {
  all <- readRDS("../data/all-sigs.rds")
}

@@

## Problem statement and motivation

\hh{'considered to be unique' by AFTE - that is right, but we should not use this argument, and I don't think we have to. It's better to focus on the problem at hand: striation marks are used to determine same source for bullets. XXX we could include some figures here XXX Current standards ask for a visual inspection of the marks under a comparison microscope by a firearms examiner. One of the problems raised by the NAS report in 2009 is the subjectivity involved in this comparison and the need for determining error rates \citep{NAS:2009}. There are different approaches used to achieve more objectivity and assess error rates. XXX here we need some examples of approaches XXX
For toolmarks, the Chumbley score has been suggested to determine same-source. XXX summary of results from original paper and the adjustment XXX
In this paper we are interested in how this method has to be adjusted to best be applied to the smaller marks on bullets. }

\begin{figure}
\centering
\includegraphics[width=\textwidth]{images/B6-B2-L6.png}
```{r, echo=FALSE, fig.width=8, fig.height=3, out.width='\\textwidth'}
land <- read.csv("../data/b6b2l6.csv")
land %>% ggplot(aes(x = y, y=resid)) +geom_line() +theme_bw()
```

\caption{\label{fig:rgl} Image of a bullet land from a confocal light microscope at 20 fold magnification.}

\end{figure}

@@

Compairing pairs of toolmarks with the intention of matching it to a tool has been studied in the past, and \citet{chumbley} have described in their paper an algorithm and analytic method that compares two toolmarks and comes to the conclusion if they are from the same tool. 
\hh{XXX anything to do with motivation has to come first - move this to the first paragraph of the intro XXX}
The motivation for this lies in the idea of developing analytical methods for which error rates can be found. This moves reduces subject bias and empirical based studies to designate two toolmarks as matches or non-matches, although \citet{chumbley} use this kind of a setup to validate their proposed analytic and quantitative algorithm.

\hh{XXX data and results come after the intro to chumbley.}
The data that is used by \citet{chumbley} is generated by a surface profilometer that gives the height in terms of distance along a linear trace. This is taken perpendicular to the striations present in the toolmark. Two such trace are then compared using the algorithm.

\hh{XXX start with defining the data, i.e. we are working with random processes $x(t)$ and $y(t)$ ... you should also go into the coarseness parameter here - that is a pre-processing step before the optimization. }
Th algoritm works in two phases, namely, an optimization step and a validation step, at the end of which a Mann Whitney U statistic is calculated. \citet{hadler} in their paper proposed an improvement to this algorithm by trying to remove mutual dependence of parameters (due to serial correlation in surface depth values of a toolmark and because of a random sampling sub step in the validation phase which makes a group of pixels to be chosen more than once and hence introduces lack of independence) in certain steps, especially because the Mann-Whitney U statistic that is later calculated in the algorithm and used as a measure to differentiate between matches and non-matches works under the assumption of independence of parameters.

@@

A maximum correlation statistic is used to identify the region of best agreement, with the maximum usually seen being near 1 for both cases of matches ( which is intuitively what we expect) and non matches (which is not what we intuitively expect) \hh{XXX get rid of the parentheses - put the content into a sentence after your first statement.}

\hh{XXX the process you describe below is the cross-correlation of two series $x(t)$ and $y(t)$. }
First, there are a very large number of correlations calculated, for eg if the window of optimization was defined as 200 and toolmark pixel length is around 1000 then we have 1200-(200-1) = 1001. Can be thought of ordered arrangements without replacement where n = 1001 and r = 1.

@@

\hh{XXX discussion of limitations should come after introducing the method. }
This introduces a problem as if we go too small in the window of optimization the unique feature of the trace segments are lost and seem similar, and too large sizes vastly reduces the weight of small features that would otherwise uniquely classify a signature and hence identify the region of agreement.

@@

```{r win-comparison, fig.width=8, fig.height=5, out.width='\\textwidth', fig.cap=" The two plots on the left show how the same shift behaves in case of a matching pair and the two plots on the right show how the different shift behaves in case of a matching pair."}
a<- matrix(NA, ncol = 3, nrow = 2)
toolmark.falsepositive<- data.frame(a)
colnames(toolmark.falsepositive)<- c("Classification", "Match", "Non-Match")
toolmark.falsepositive$Classification<- c("Match", "Non-Match")
toolmark.falsepositive$Match<- c(47,0)
toolmark.falsepositive$`Non-Match`<- c(3,50)
toolmark.falsepositive

@@

@@

```{r type2, fig.width=8, fig.height=5, out.width='\\textwidth', fig.cap="Type II error rates observed across a range of window sizes for optimization $wo$. For a window size of $wo = 120$ we see a drop in type II error rate across all type I rates considered. Smaller validation sizes $wv$ are typically associated with a smaller type II error."}
errors %>% filter(wv %in% c(30,50)) %>%
  ggplot(aes(x = wo, y = beta, colour=factor(alpha))) + 
  geom_point(aes(shape=factor(wv))) +
#  facet_grid(.~wo, labeller="label_both") + 
  geom_smooth(aes(group=alpha), se=FALSE, method="loess") +
  theme_bw() +
  scale_colour_brewer(expression("Nominal\ntype I error"~alpha), palette="Set2") +
  scale_shape_discrete("Size of validation\nwindow, wv") +
  xlab(expression("Window size for optimization, "~wo)) +
  ylab("Type II error rate")
